---
title: "Working with XML and JSON in R"
author: "Esteban Aramayo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: false
    #number_sections: true
    toc_float: true
    collapsed: false
    theme: united
    highlight: tango
# output: rmarkdown::html_vignette
# vignette: >
#   %\VignetteIndexEntry{Vignette Title}
#   %\VignetteEngine{knitr::rmarkdown}
#   %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, tidy=TRUE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r load-packages, message=FALSE}
# Packages to be used
library(tidyverse)
library(rvest)   # used to manipulate HTML 
library(XML)
library(xml2)
library(jsonlite)
```


## Problem Overview

Pick three of your favorite books on one of your favorite subjects. At least one of the books should have more than one author. For each book, include the title, authors, and two or three other attributes that you find interesting.
Take the information that you’ve selected about these three books, and separately create three files which store the book’s information in HTML (using an html table), XML, and JSON formats (e.g. “books.html”, “books.xml”, and “books.json”). To help you better understand the different file structures, I’d prefer that you create each of these files “by hand” unless you’re already very comfortable with the file formats.
Write R code, using your packages of choice, to load the information from each of the three sources into separate R data frames. Are the three data frames identical?


## Read the data from HTML, XML, and JSON source files


### Read book data from an HTML file

Read book data from an HTML file and store it in a dataframe

```{r read-html, message=FALSE}
books_html <- rvest::read_html("Books.html")
df_html <- books_html %>%
  html_element("table") %>%
  html_table()
```



### Read book data from an XML file

Read book data from an XML file and store it in a dataframe

```{r read-xml, message=FALSE}
books_xml <- XML::xmlParse("Books.xml")
root_xml <- XML::xmlRoot(books_xml)
df_xml <- xmlToDataFrame(root_xml)

```




### Read book data from a JSON file

Read book data from an JSON file and store it in a dataframe

```{r read-json, message=FALSE}
books_json <- jsonlite::read_json("Books.json", simplifyVector = TRUE)
df_json <- books_json$Books
```




## Compare the data type of the 3 resulting variables holding the content of the read files

**The resulting data types for HTML, XML and JSON sources are different.**


```{r, echo=TRUE, results='asis'}
class(books_html)
class(books_xml)
class(books_json)
```




## Compare the Output of the 3 resulting dataframes

The resulting content of the dataframes for all three sources are different.


**Dataframe output from HTML source:** 

The authors columns are repeated for each author and are filled with "NA" when fewer authors exist

```{r, echo=FALSE, results='asis'}
knitr::kable(head(df_html, 10))
```

**Dataframe output from XML source:**

The authors names are combined in one column but they are separated by single spaces

```{r, echo=FALSE, results='asis'}
knitr::kable(head(df_xml, 10))
```

**Dataframe output from JSON source:**

The authors names are combined in one column but they are separated by commas

```{r, echo=FALSE, results='asis'}
knitr::kable(head(df_json, 10))
```


## Compare the data type of the 3 resulting dataframes

The resulting data types for XML and JSON are the same, but the HTML is not.


```{r, echo=TRUE, results='asis'}
class(df_html)
class(df_xml)
class(df_json)
```







