---
title: "Week Three - R Character Manipulation and Date Processing"
author: "Esteban Aramayo"
#output: github_document
output:
    html_document:
        toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE)
```

```{r echo=FALSE, include=FALSE}
library(readr)
library(tidyverse)
library(stringr)
```

### 1. Overview

Please deliver links to an R Markdown file (in GitHub and rpubs.com) with solutions to the problems below.  You may work in a small group, but please submit separately with names of all group participants in your submission.



### 2. Filtering data using Regular Expressions (regex)

Using the 173 majors listed in fivethirtyeight.com’s College Majors dataset (<A href="https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/">The Economic Guide To Picking A College Major</a>), provide code that identifies the majors that contain either "DATA" or "STATISTICS".


```{r read-538-college-majors, message=FALSE}

# define the URL for the raw csv file containing major list
majors_url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/majors-list.csv"

# read csv file and store the records into a data frame
majors <- read_csv(majors_url)

# convert the data frame to a tibble
mt <- as_tibble(majors)

# from the list of majors filter only those which contain the words DATA OR STATISTICS
tbl2 <- mt %>% filter(grepl('DATA|STATISTICS', Major, ignore.case = TRUE))

# count the total of major found in the majors list
total_majors <- nrow(majors) # Actual total of majors in data set

#An actual total of `total_majors` majors were found in the current site's data set.


# display the list of filtered majors
knitr::kable(tbl2, caption = 'Table: List of Majors that contain the words "DATA" or "STATISTICS"')


```







### 3 Data Transformation

Write code that transforms the data below:

[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"

[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  

[9] "elderberry"   "lime"         "lychee"       "mulberry"    

[13] "olive"        "salal berry"

Into a format like this:

c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")



```{r}

# stored the character data as is, including spaces, carriage returns, double quotes
raw_data <- '[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"

[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  

[9] "elderberry"   "lime"         "lychee"       "mulberry"    

[13] "olive"        "salal berry"'


# define a regex to match words surrounded by double quotes "\b\w+\s?\w*\b"
regex_double_quote_surrounded_words <- '\\b\\w+\\s?\\w*\\b"'

# extract the quoted words from the string and store them into a list
extracted_quoted_words <- str_extract_all(raw_data, regex_double_quote_surrounded_words)

# concatenate the quoted words and store them in a character vector as per the ask
new_character_format <- paste(extracted_quoted_words, ",")


extracted_quoted_words

new_character_format

```






### 4 Additional exercises


```{r}
# Build a tibble with a list of sample words to be tested against the regular expressions
dt_test <- tibble( words = c("Ohio","Pennelope","Kabbak","tomato","eleven","church","Pop","papa","Robocop","look","Pajama","noN","XXXrated"))
```


```{r, results=TRUE, echo=FALSE}
knitr::kable(dt_test, caption = 'Table: List of Sample words to be used to test the Regular Expressions')
```


The two exercises below are taken from R for Data Science, 14.3.5.1 in the on-line version:


#### 4.1 Interpreting regular expresions



Describe, in words, what these expressions will match:

(.)\1\1

It matches a character that repeats 3 times consecutively.

```{r}
dt_matches <- dt_test %>% dplyr::filter(str_detect(words, regex('(.)\\1\\1', ignore_case = T)))

knitr::kable(dt_matches)

```



"(.)(.)\\2\\1"

It matches two characters followed by the same two characters but in reversed order.

```{r}
dt_matches <- dt_test %>% dplyr::filter(str_detect(words, regex("(.)(.)\\2\\1", ignore_case = T)))

knitr::kable(dt_matches)

```



(..)\1

It matches two characters followed by the same two characters.

```{r}
dt_matches <- dt_test %>% dplyr::filter(str_detect(words, regex("(..)\\1", ignore_case = T)))

knitr::kable(dt_matches)

```






"(.).\\1.\\1"

It matches a character repeated 3 times in non-consecutive positions.

```{r}
dt_matches <- dt_test %>% dplyr::filter(str_detect(words, regex("(.).\\1.\\1", ignore_case = T)))

knitr::kable(dt_matches)

```




"(.)(.)(.).*\\3\\2\\1"


It matches 3 characters followed by the same 3 characters but in reversed order.

```{r}
dt_matches <- dt_test %>% dplyr::filter(str_detect(words, regex("(.)(.)(.).*\\3\\2\\1", ignore_case = T)))

knitr::kable(dt_matches)

```






#### 4.2 Constructing regular expressions


#### 4.2.1 Regular expression to match words that start and end with the same character.
```{r}

# define a regular expression that will match words starting and ending with the same character: \b(\w)\w*\1\b
regex_same_1st_n_last_char <- "\\b(\\w)\\w*\\1\\b"

# filter the words that match the regular expression
dt_flc_filtered <- dt_test %>% dplyr::filter(str_detect(words, regex(regex_same_1st_n_last_char, ignore_case = T)))

knitr::kable(dt_flc_filtered, caption = 'Table: List of words starting and ending with the same character')

```



#### 4.2.3 Regular expression to match words that contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)
```{r}

# define a regular expression that will match words that contain a repeated pair of letters:  \b(..)\w*\1\b
regex_2_rep_letters <- "\\b(..)\\w*\\1\\b"

# filter the words that match the regular expression
dt_flc_filtered <- dt_test %>% dplyr::filter(str_detect(words, regex(regex_2_rep_letters, ignore_case = T)))

knitr::kable(dt_flc_filtered, caption = 'Table: List of words that contain a repeated pair of letters')

```



#### 4.2.4 Regular expression to match words that contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)
```{r}

# define a regular expression that will match words that contain one letter repeated in at least three places:  ([a-z])\w*\1\w*\1
regex_3_rep_letters <- "([a-z])\\w*\\1\\w*\\1"

# filter the words that match the regular expression
dt_flc_filtered <- dt_test %>% dplyr::filter(str_detect(words, regex(regex_3_rep_letters, ignore_case = T)))

knitr::kable(dt_flc_filtered, caption = 'Table: List of words that contain one letter repeated in at least three places')

```




