---
title: "Project 01 - Data Analysis"
author: "ESTEBAN ARAMAYO"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: true
    collapsed: false
    theme: united
    highlight: tango
---

<H2>Overview</H2>

In this project, we are given a text file with chess tournament results where the information has some structure. Our job is to create an R Markdown file that generates a .CSV file (that could for example be imported into a SQL database) with the following information for all of the players:

Player’s Name, Player’s State, Total Number of Points, Player’s Pre-Rating, and Average Pre Chess Rating of Opponents

For the first player, the information would be:
Gary Hua, ON, 6.0, 1794, 1605

1605 was calculated by using the pre-tournament opponents’ ratings of 1436, 1563, 1600, 1610, 1649, 1663, 1716, and dividing by the total number of games played.


```{r load-packages, message=FALSE}
# Libraries to be used
library(tidyverse)
library(data.table)
```



<H2>Read Chess Tournament Source Data</H2>
```{r read-src-data}

tourney_file_name <- "C:\\tmp\\tournamentinfo.txt"

# Read the data from the tournament file and store it in a data.table (skip the first 4 rows containing unnecessary header data)
mdt <- data.table(read_csv(tourney_file_name, col_types =  list(col_character()), col_names = c("data"),  skip = 4))


```

Let's peak into the data read from the file.
```{r}
# Take a peak at the data from the table
head(mdt,10)
```



<h2>Parsing out the data</h2>

Extract the lines of text corresponding to the players and their ratings.

```{r parse-out-text-lines}

# From the table, extract the text rows that correspond to the players (skip the dashed lines)
players <- mdt[seq(1, nrow(mdt), 3)]

# From the table, extract the text rows that correspond to the players' ratings (skip the dashed lines)
player_ratings <- mdt[seq(2, nrow(mdt), 3)]

```

Show the results of the extracted lines of text.

```{r}
# take a peak at the players data lines
head(players, 10)

# take a peak at the players' ratings data lines
head(player_ratings, 10)
```

Using regular expressions, parse out the values from each line of text.

```{r message=FALSE, warning=FALSE}
# Extract the values from the players table
Pair_Num <- as.integer(str_extract(players$data, "\\d+"))                      # 1st sequence of numeric characters
Player_Name <- str_trim(str_extract(players$data, "(\\w+\\s){2,3}"))           # 1st instance of 2 or 3 words followed by white space
Total_Points <- as.numeric(str_extract(players$data, "\\d+\\.\\d+"))           # 2nd sequence of numeric characters
Opponents <- str_extract_all(str_extract_all(players$data, "\\d+\\|"), "\\d+") # All instances of numeric characters followed by a pipe |

# Extract the values from the player ratings table
Player_State <- str_extract(player_ratings$data, "\\w+")              # 1st word on the line
Pre_Rating <- as.integer(str_extract(str_extract(player_ratings$data, "[^\\d]\\d{3,4}[^\\d]"), "\\d+")) # 1st instance of a number of at least 3 but no more than 4 digits 
Post_Rating <- as.integer(str_extract(str_extract(player_ratings$data, "->\\s*\\d{3,4}"), "\\d+")) # 1st instance of a number of at least 3 but no more than 4 digits, which come after "-> ".

```

<h2>Calculate the Arpad Elo's mean rating</h2>

Calculate the Arpad Elo's mean rating for each player's opponents.

For this, we traverse the vector of Pre-Ratings of each player and get the pre-ratings of their opponents. Because the opponents' IDs (Pair number) are stored in a vector inside a vector, we need to flatten it before we can use the values from the inner vector. Using those IDs, we look up the pre-rating corresponding to each of those IDs (Pair number) and then calculate the mean of all of them. We round the mean to the nearest point (no decimal digits). 

```{r}
# Calculate the Arpad Elo's mean rating for each player's opponents
total_Players = nrow(players)

Avg_Opponent_Pre_Rating <- total_Players

for (i in 1:total_Players) { 
  # for each player
  Avg_Opponent_Pre_Rating[i] <- round(
    mean(
      Pre_Rating[as.numeric(
        flatten(
          Opponents[Pair_Num[i]]
        )
        )]
    ), digits = 0
  ) 
}
```


<h2>Build a table with all the final results</h2>

```{r}

# build a table to be used as the source for the CSV file
DT = data.table(
  Player_Name,
  Player_State,
  Total_Points,
  Pre_Rating,
  Avg_Opponent_Pre_Rating
)

```

```{r echo=FALSE}
knitr::kable(DT[1:10, ], caption = 'Final results table')
```



<h2>Export the results to a CSV file </h2>
```{r}
# Write the data to CSV file
write_csv(
  x = DT,
  file = "c:\\tmp\\chess_tournament_player_output.csv",
  quote_escape = "double"
)
```

<H2>Player scored the most points relative expected result</H2>
```{r}

# Which player scored the most points relative to his or her expected result?
Point_Diff <- Post_Rating - Pre_Rating

elo_rating = data.table (
  Player_Name,
  Pre_Rating,
  Post_Rating,
  Point_Diff,
  Avg_Opponent_Pre_Rating
)


elo_rating <- elo_rating[order(-Point_Diff)]

```

```{r echo=FALSE}
knitr::kable(elo_rating, caption = 'ELO ratings')
```

PLayer with the highest difference between his/her Pre-rating and Post-rating


```{r echo=FALSE}
knitr::kable(elo_rating[1], caption = 'ELO ratings')
```







